Put Your Money Where Your Mouth Is: Prediction Markets and Climate Change

### 2016-03-03 paper.Rmd
paper work. moved more util funcs to be sourced too.

---
title: "Convergence of Forecasting Models in Climate Prediction Markets: An Agent-Based Simulation"
author: "John J. Nay, Martin Van der Linden, Jonathan Gilligan"
date: "March 2, 2016"
output: 
  pdf_document:
    keep_tex: true
bibliography: library.bib
---

```{r fig_ref, echo=FALSE}
figRef <- local({
  tag <- numeric()
  created <- logical()
  used <- logical()
  function(label, caption, prefix="Fig.", sep=".",
      prefix.highlight="**") {
    i <- which(names(tag) == label)
    if(length(i) == 0){
      i <- length(tag) + 1
      tag <<- c(tag, i)
      names(tag)[length(tag)] <<- label
      used <<- c(used, FALSE)
      names(used)[length(used)] <<- label
      created <<- c(created, FALSE)
      names(created)[length(created)] <<- label
    }
    if(!missing(caption)){
      created[label] <<- TRUE
      paste0(prefix.highlight, prefix, " ", i, sep, prefix.highlight, " ", caption)
    } else {
      used[label] <<- TRUE
      paste(prefix, tag[label])
    }
  }
})
# adapted from a function on this page: http://galahad.well.ox.ac.uk/repro/
rinline <- function(code){
  sprintf('`r %s`', code)
}
```

```{r, caption_setting, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(knitr)
knit_hooks$set(htmlcap = function(before, options, envir) {
  if(!before) {
    paste('<p class="caption">',options$htmlcap,"</p>",sep="")
    }
    })
# small margins
knit_hooks$set(small_mar = function(before, options, envir) {
    if (before) par(mar = rep(0.0001, 4))  # smaller margins all around
})
# opts_chunk$set(dev="tiff", 
#                dev.args=list(compression="lzw", type="cairo"),
#                dpi=300)
```

```{r tab_ref, echo=FALSE}
tabRef <- local({
    tag <- numeric()
    created <- logical()
    used <- logical()
    function(label, caption, prefix = "Table", 
        sep = ".", prefix.highlight = "**") {
        i <- which(names(tag) == label)
        if (length(i) == 0) {
            i <- length(tag) + 1
            tag <<- c(tag, i)
            names(tag)[length(tag)] <<- label
            used <<- c(used, FALSE)
            names(used)[length(used)] <<- label
            created <<- c(created, FALSE)
            names(created)[length(created)] <<- label
        }
        if (!missing(caption)) {
            created[label] <<- TRUE
            paste0(prefix.highlight, prefix, " ", i, sep, prefix.highlight, 
                " ", caption)
        } else {
            used[label] <<- TRUE
            paste(prefix, tag[label])
        }
    }
})
```

```{r multi_plot, echo=FALSE}
multi_plot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

library(ggplot2)
theme_set(theme_classic())
library(pander)
```

@@

We present an agent-based simulation of a climate prediction market. Traders adapt their beliefs about the future global temperatures based on the monetary performance of their neighboring traders in a social network. We conduct simulation experiments with the model to determine how a variety of market factors and the climate data-generating process affect whether climate prediction markets foster a convergence of market participants' beliefs regarding a true model of the global climate.

@@

Attempts to foster such consensus face many social and psychological challenges, some of which could be adressed by creating climate prediction markets [@Hsu2011; @Vandenbergh2013f]. The idea of using prediction markets to efficiently aggregate information about uncertain events has been widely discussed [@Horn2014c]. Prediction markets have been shown to have interesting theoretical properties [@Set1998; @Hanson2012] and to perform well in terms of prediction accuracy and information aggregation in experiments [@Hanson2006; @Healy2010], agent-based models (ABM) [@Klingert2012c; @Jumadinova2011a], and the real-world [@Wolfers2006]. We simulate a prediction market where traders exchange securities related to climate outcomes to explore whether, and under what market and climate conditions, prediction markets may be useful for increasing convergence of climate beliefs.

@@

@Tseng2010a study an ABM of a continuous double auction market with various types of market strategies (including two variants of the zero-intelligence agent). They compare the behavior of their simulation with data from a prediction market for the outcome of political elections. They find that despite their simplicity, zero-intelligence agents capture some salient features of real market data. @Klingert2012c reach similar conclusions with respect to experimental data from @Hanson2006. In their simulation, @Klingert2012c compare continuous double auctions with logarithmic market scoring rules and find that the latter generally outperform continuous double auctions in terms of prediction accuracy. In some variants of the two last models, agents' strategies feature learning, but these are rather basic and unrelated to a formal predictive model.

@Jumadinova2011a conducted research more similar to ours, as they explicitly include information about uncertain events in their model. @Jumadinova2011a study a continuous double auction model in which agents update their beliefs about the uncertain events based on information on the event. New beliefs are modeled as a weighted sum of the last period's beliefs and market prices. The agents' information set changes every period. The better the information set, the more likely the agent is to put higher weights on last period's prices when revising their beliefs. Not much structure is imposed on the information set and the way it is used to generate the weights in the update function. Therefore, the model does not allow for investigating the convergence of predictive models. @Ontanon2009 study the effect of deliberation on a simulated prediction market. They assume that agents use Case-Based Reasoning [@Aamodt1994] to reason and debate about uncertain outcomes with their neighbors in a social network.  Although the model could in theory be used to assess convergence of predictive models by looking at the arguments agents adopt after the deliberation, this is not done by the authors. Also, the way beliefs are formed and revised through argumentation is abstract and not ideal for studying climate prediction markets.

@@

From a public policy perspective, changing the explanatory models of market participants is one of the most important roles that prediction markets might play. This is perhaps the most important social benefit of prediction markets given the comparable predictive power of statistical models, which are less expensive than setting up and maintaining prediction markets [@goel_prediction_2010]. Effective climate policy not only requires an accurate consensus on future climate *outcomes*. It also requires an accurate consensus on the causal *mechanisms* influencing such outcomes. If people agree that temperature will rise, but some believe it will be due to greenhouse gases, while others believe that it will be caused by increased solar activity, inconsistent and ineffective policies may be implemented. Our model is designed to investigate whether prediction markets can foster such convergence in predictive models.

@@

JG - describe the temperature time series models here.

@@

Traders use one of the two models to forecast future temperature anomalies. These models are interpreted as the trader's beliefs about the true climate process, the driving factor of long-term global temperatures. They represent pervasive positions on climate change in the public debate. In order to approximatively match the current configuration of beliefs in climate change in the United States, model 1 is randomly assigned to half of the traders, while 2 is assigned to the other half of the traders.

@@

## Time

@@

The sensitivity analysis is based on a Latin Hypercube Sampling of parameter sets from the following distributions [@beachkofski_improved_2002,@carnell_lhs_2012].

@@

Our sensitivity analysis, `r figRef("sa")`, shows that

```{r sa, eval=TRUE, echo=FALSE, message=FALSE, htmlcap=figRef("sa", "Estimated Effects on Convergence of Beliefs in Future Scenario "), fig.width=4, fig.height=4.5}
# library(eat)
# load("output/src_future.Rda")
# plot(src_future, outcome_var = "Convergence") + theme_classic() + theme(text=element_text(family='Palatino'))

@@

```{r time, eval=TRUE, echo=FALSE, message=FALSE, htmlcap=figRef("time", "Time"), fig.width=5, fig.height=10}
source("utils/time_plot.R")
time_plot() + theme(text=element_text(family='Palatino'))
```

>>>>

---
title: "Put Your Money Where Your Mouth Is: Prediction Markets and Climate Change"
author: "John J. Nay, Martin Van der Linden, Jonathan Gilligan"
date: "March 3, 2016"
output: 
  pdf_document:
    keep_tex: true
bibliography: library.bib
---
```{r fig_ref, echo=FALSE}
source("utils/fig_ref.R") # tabRef and figRef functions
source("utils/multi_plot.R")
``` 
```{r, caption_setting, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(knitr)
knit_hooks$set(htmlcap = function(before, options, envir) {
  if(!before) {
    paste('<p class="caption">',options$htmlcap,"</p>",sep="")
    }
    })
# small margins
knit_hooks$set(small_mar = function(before, options, envir) {
    if (before) par(mar = rep(0.0001, 4))  # smaller margins all around
})
# opts_chunk$set(dev="tiff", 
#                dev.args=list(compression="lzw", type="cairo"),
#                dpi=300)
```
```{r multi_plot, echo=FALSE}
source("utils/multi_plot.R")
library(ggplot2)
theme_set(theme_classic())
library(pander)
```

@@

We present a simulation of a climate prediction market. Traders adapt their beliefs about the future global temperatures based on the monetary performance of their neighboring traders in a social network. We conduct experiments with the model to determine how a variety of market factors and the climate data-generating process affect whether climate prediction markets foster a convergence of market participants' beliefs regarding a true model of the global climate.

@@

Attempts to foster such consensus face many social and psychological challenges, some of which could be adressed by creating climate prediction markets where participants can "put their money where their mouths are" [@Hsu2011; @Vandenbergh2013f]. The idea of using prediction markets to efficiently aggregate information about uncertain events has been widely discussed [@Horn2014c]. Prediction markets have been shown to have interesting theoretical properties [@Set1998; @Hanson2012] and to perform well in terms of prediction accuracy and information aggregation in experiments [@Hanson2006; @Healy2010], simulation models [@Klingert2012c; @Jumadinova2011a], and the real-world [@Wolfers2006]. We simulate a prediction market where traders exchange securities related to climate outcomes to explore whether, and under what market and climate conditions, prediction markets may be useful for increasing convergence of climate beliefs.

@@

@Tseng2010a create an agent-based model (ABM) of a continuous double auction market with various types of market strategies, including two variants of the zero-intelligence agent. They compare the behavior of their simulation with data from a prediction market for the outcome of political elections. They find that despite their simplicity, zero-intelligence agents capture some salient features of real market data. @Klingert2012c reach similar conclusions with respect to experimental data from @Hanson2006. In their simulation, @Klingert2012c compare continuous double auctions with logarithmic market scoring rules and find that the latter generally outperform continuous double auctions in terms of prediction accuracy. In some variants of the two last models, agents' strategies feature learning, but these are rather basic and unrelated to a formal predictive model.

@Jumadinova2011a conducted research more similar to ours, as they explicitly include information about uncertain events in their model. @Jumadinova2011a study a continuous double auction model in which agents update their beliefs about the uncertain events based on information on the event. New beliefs are modeled as a weighted sum of the last period's beliefs and market prices. The agents' information set changes every period. The better the information set, the more likely the agent is to put higher weight on last period's prices when revising their beliefs. Little structure is imposed on the information set and the way it is used to generate the weights in the update function. Therefore, the model does not allow for investigating the convergence of predictive models. @Ontanon2009 study the effect of deliberation on a simulated prediction market. They assume that agents use Case-Based Reasoning [@Aamodt1994] to reason and debate about uncertain outcomes with their neighbors in a social network. Although the model could in theory be used to assess convergence of predictive models by looking at the arguments agents adopt after the deliberation, this is not done by the authors. Also, the way beliefs are formed and revised through argumentation is too abstract for studying climate prediction markets, where beliefs should be based on statistical time series models.

@@

From a public policy perspective, changing the explanatory models of market participants is one of the most important roles that prediction markets might play. This is perhaps the most important social benefit of prediction markets given the comparable predictive power of statistical supervised learning models, which can provide information at a much lower cost than creating and maintaining a prediction market [@goel_prediction_2010]. Effective climate policy not only requires an accurate consensus on future climate *outcomes*. It also requires an accurate consensus on the causal *mechanisms* influencing such outcomes. If people agree that temperature will rise, but some believe it will be due to greenhouse gases, while others believe that it will be caused by increased solar activity, inconsistent and ineffective policies may be implemented.

@@

We start with the historical temperature record and project future climates under two alternative theories:

Temperature proportional to log(CO2) or Total Solar Irradiance (conventional science vs. popular alternative among doubters). Add AR(1) noise (best ARMA noise model for historical record).

JG - describe the temperature time series models here. Add the plots of the time series diverging thats on middle of poster. Add citations for the models to library.bib.

@@

Traders use one of the two models (temperature depends on CO2 or TSI) to forecast future temperature anomalies. These models are interpreted as the trader's beliefs about the true climate process, the driving factor of long-term global temperatures. They represent pervasive positions on climate change in the public debate. In order to approximatively match the current configuration of beliefs in climate change in the United States, model 1 is randomly assigned to half of the traders, while 2 is assigned to the other half of the traders.

@@

## Model Dynamics

Bet on temperature six years in future.
During six-year period, traders buy and sell futures.
Every year: Bayesian updating of traders personal predictions for future temperature based on current year’s temperature.
At end of six-year period, winners collect money.
Traders revise beliefs about climate models, based on ideology and beliefs of top earners in their social network.

@@

## Historical Data

As a validation that the market is operating correctly, we run the market with actual temperatures 1880 to 2015 and market betting from 1931 to 2014. As warming signal begins to show up in 1980s, traders begin to converge toward log(CO2) model.

## Future Scenarios

The sensitivity analysis is based on a Latin Hypercube Sampling of parameter sets from the following distributions [@beachkofski_improved_2002; @carnell_lhs_2012].

@@

Our sensitivity analysis, `r figRef("sa")`, shows that ideology, risk tolerance, the number of traders, and the segmentation of the social network significantly affect convergence. Segmented social networks can create an "echo-chamber" effect.

```{r sa, eval=TRUE, echo=FALSE, message=FALSE, htmlcap=figRef("sa", "Estimated effects of model parameters on convergence of beliefs in future scenario."), fig.width=3.8, fig.height=4.5}
# library(eat)
# load("output/src_future.Rda")
# plot(src_future, outcome_var = "Convergence") + theme_classic() + theme(text=element_text(family='Palatino'))

@@

```{r time, eval=TRUE, echo=FALSE, message=FALSE, htmlcap=figRef("time", "Time"), fig.width=5, fig.height=9}
source("utils/time_plot.R")
time_plot() + theme(text=element_text(family='Palatino'))
```
